---
title: "Multivariate Statistics - Assignment 2 "
author:
  - Elena 
  - Damiano
  - Xierui
  - Aharon
date: "12/1/2020"
output:
  pdf_document: 
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

#where I can find my data sources
base_url <- '~/GitHub/kul-multivariate-a2/src/'
#list.files('./src/', all.files=FALSE, full.names = FALSE, recursive = TRUE)
```

\newpage

# Assignment 2 Multivariate Statistics 

This document ...

# Task 1 
```{r Task_1_setup, include=FALSE}
#Load dwvs
library(candisc)
library(car)
load(paste0(base_url,"dwvs.Rdata"))
```


## Introduction 
  In the present part of the report, we will investigate to what extent we will be able to classify respondents in their country, and then we will compare the performance of different classifiers.

## Data 
  The data have been obtained from the 6th Wave of the World Value Survey, which was carried out between 2010 and 2013. The data include the standardized scores of 3929 respondents of 3 countries on 32 variables, that have been summarized with 7 factors obtained using exploratory factor analysis with oblique rotation. 
The 7 factors related to the 32 variables are: 


1.	**Rights**, that it’s related to homosexuality, prostitution, abortion, divorce, sex before marriage, suicide;
2.	**Steal** that it’s related to claiming benefits, avoiding fare, stealing property, cheating taxes, accept a bribe;
3.	**Crime** that it’s related to robberies, alcohol, police-military, racist behavior, drug sale;
4.	**Religion** that it’s related to: attend religious services, pray, the importance of God;
5.	**Realize self** that it’s related to creative, rich, spoil oneself, be successful, exciting life;
6.	**Do good** that it’s related to security, do good, behave properly, protect environment, tradition;
7.	**Violence** that it’s related to beat wife, parents beating children, violence.


## Methodology

  To investigate the possibility to classify the respondents in their country based on the 7 factors we have used the canonical discriminant analysis. We have applied the linear regression function with 7 predictors and 1 dependent variable, the Country. Then to the output, we have applied the Canonical Discriminant Analysis.


```{r Task_1_1}

lm.out<-lm(cbind(F_rights, F_steal, F_crime,F_religion,F_realizeself,F_dogood ,F_violence)~as.factor(country), data=dwvs)
candisc.out<-candisc(lm.out)
print (candisc.out)


```

As we can see both the Square Canonical Correlation are significant, but the discriminating power to separate between the groups is higher for the first than for the second discriminant function: 0.81 and 0.38 respectively.
The LR test indicates that the discriminant analysis is meaningful. The first test’s null hypothesis is H0: λ1=λ2=0 and this hypothesis as we can see from the P-value is rejected. The null hypothesis of the first test it’s equivalent to the test for H0: μNetherlands=μNigeria=μPhilippines.
The second LR test indicates that **H0:λ2=0**, and also this null hypothesis is rejected. So even if the second discriminant function has less discriminant power cannot be omitted and it’s statistically meaningful. 
On our analysis, we have also applied two different tests for centroids and to test the equal covariance. 
To see if the three-country has different centroids and confirm the results of the canonical discriminant analysis we have applied on the linear regression the function Manova:

```{r Task_1_2}
summary(Manova(lm.out), test="Wilks")

```

The p-value is small and the test confirms that the analysis is meaningful and that at least there is a pair of centroids that differs significantly. The function Manova in r doing the Wilks Lambda test uses the Rao approximation. 
To test the assumption on equal population covariance we have applied to the linear regression the function boxM: 


```{r Task_1_3}
boxM(lm.out)

```

### Plot

To complete the Canonical Discriminant analysis we have plotted the three countries and the 7 variables. 

```{r Task_1_4}
plot(candisc.out,col=c("red","blue","black"),pch=c(1,2,3),cex=1.2)

```

We can see that the group of individuals in red are Netherland citizens, the group of individuals in green are Nigerians citizens and the group of individuals in black are Philippines citizens. In blue we can see the 7 explanatory variables. The plot shows a clear separation between Netherland and the other two countries on the first discriminant function while the second discriminant function could help to separate Nigeria and the Philippines. 
The first discriminant function especially correlates with the factors: rights, religion, realize self, and do good; whereas the second discriminant function correlates with the factors: steal and realize self. The two-factor crime and violence has a lower correlation on the two factors and so has in this analysis lower importance on separate the 3 countries.

The test of Box indicates that H0 of equal covariance matrices across groups is not supported by data. So the Linear discriminant Analysis will not work correctly in that case. 

\newpage
# Task 2
```{r Task_2_setup, include=FALSE}

library(tree)
library(randomForest)
library(gbm)
library(pdp)
library("MASS")
library(heplots)
library(candisc)
load(paste0(base_url,"spamdata.Rdata"))
```
## Introduction

  For task 2, we are going to deal with a dataset containing information about 4601 webmails. We have 48 variables describing the frequency of some specific words like “remove” in each observation, 6 variables describing the frequency of some specific chars like “$” in one observation, and three variables, **capital_run_length_longest**, **capital_run_length_average** and **capital_run_length_total**, describing the length of the longest uninterrupted sequence of capital letters, the average length of uninterrupted sequences of capital letters, and the total number of capital letters in each observation respectively. We also have a variable called spam, which indicates whether this webmail is a spam with 0 and 1, where 1 for spam, and 0 for not spam. Here all our variables are numeric type.

Our task is to use these 57 attribute variables to classify whether a webmail is spam.


## Methodology

  In order to validate the accuracy of our methods, we firstly divide our dataset into a train set, which contains 2500 observations, and a test set, which contains 2101 observations. We use the train set to train our models, and then apply it to the test set to validate it’s accuracy.


## Results

### Classification Trees
  In this part, we are going to discuss the results obtained by complex tree model and pruned tree model.
  
  We begin with construct a complex tree model by dividing our observation into small non-overlapping regions according to some numerical criteria. Here we split our dataset until each leaf of our classification tree contains only less then 2 observations. The method used here is recursive binary splitting.
```{r Task_2_1, include=FALSE}

spamdata$spam<-factor(ifelse(spamdata$spam==1,"1_yes","0_no"))

#create train and test set
set.seed(0829539)
trainsize<-2500
train = sample(nrow(spamdata), trainsize)
data.train<-spamdata[train,]
data.test<-spamdata[-train,]

```


```{r Task_2_2}
err<-function(observed,predicted)
{tab<-table(observed,predicted)
print(tab)
err<-1-sum(diag(tab))/sum(tab)
return(err)
}

#grow complex tree using deviance as criterion
tree.mod <-tree(spam~., data.train, control=tree.control(nobs=2500, minsize=2, mincut=1), 
                split="deviance")
summary(tree.mod)

```

### Tree Plots

```{r Task_2_3, echo=FALSE}
plot(tree.mod)
text(tree.mod, pretty=0, cex=1.4)
```

We can see clearly here that criterias concerning the frequency of “$”, “remove”,  “!”,  “hp”,  “our”,  “free”,  “edu” as well as the length of the longest uninterrupted sequence of capital letters are used for splitting. It’s actually quite reasonable, because from our own experience, spam webmails are always advertisements on money related topics or education related topics, and are always fulled with words in capital letters, together with exclamation symbols, to draw attention.

Since the process of recursive binary splitting may lead to overfitting, where we obtain a complex tree with a good fit on the training data, but with a poor performance on test data, we introduce the process of tree pruning. Actually, a smaller tree with fewer splits may have a lower variance at the cost of acceptable little bias.
In order to decide the optimal tuning parameter which leads to both much lower variance and acceptable bias, we use cross-validation to make a selection. In fact, the least cross-validation error implies the least probability of overfitting, as it’s also a train-test process.



```{r Task_2_4}
#use cross-validation to select tuning parameter for pruning the tree
set.seed(0829539)
cv.out=cv.tree(tree.mod,K=5)
par(cex=1.4)
plot(cv.out$size,cv.out$dev,type='b')

```

However, in this specific task, we can see that the cross-validation error is monotonously decreasing, so the optimal fold number is the original fold number. We choose best size=12 here such that the pruned tree model here is exactly the same as our complex tree model.

```{r Task_2_5}
#prune the tree
prune.mod=prune.tree(tree.mod,best=12)
plot(prune.mod)
text(prune.mod,pretty=0)

```

Now we validate the accuracy of our classification tree model with the test data.

```{r Task_2_6}
#make predictions on training and test set using the unpruned tree
pred.train<-predict(tree.mod,newdata=data.train)
classif.train<-ifelse(pred.train[,2]>=pred.train[,1],1,0)
err(data.train$spam,classif.train)
pred.test<-predict(tree.mod,newdata=data.test)
classif.test<-ifelse(1*pred.test[,2]>=pred.test[,1],1,0)
err(data.test$spam,classif.test)
```


```{r Task_2_7}
#make predictions on training and test set using the pruned tree
pred.train<-predict(prune.mod,newdata=data.train)
classif.train<-ifelse(pred.train[,2]>=pred.train[,1],1,0)
err(data.train$spam,classif.train)
pred.test<-predict(prune.mod,newdata=data.test)
classif.test<-ifelse(1*pred.test[,2]>=pred.test[,1],1,0)
err(data.test$spam,classif.test)
```

We can conclude that our classification model performs very well. since the complex tree is the same as pruned tree here, we can see that the test error is just very slightly higher than the train error. There is not much overfitting here.


### Different Classification Models

Now for this part, we are going to compare different classification models using two different classification scenarios.

Firstly, we consider the senario where we have ifferent prior probabilities and equal classification costs. We use the entire dataset to figure out the prior probability.

```{r Task_2_8}
#(a)Account for different prior probabilities and equal classification costs 
# (1) linear discriminant analysis
set.seed(0829539)
ldaS.out<-lda(spam~.,data=spamdata)
print(ldaS.out)

```

So we take P(is spam)=0.394, P(not spam)=0.606 as our prior probability, and we can verify that their sum equals 1.


The four classification models we are going to compare is 1) Linear Discriminant Analysis, 2) Bagging, 3) Random Forests and 4)Gradient Boosting. Since the dimension here is very high and the mathematical assumption of our dataset is very weak, we do not suppose that Linear Discriminant Analysis with out Principle Component analysis here will have very good performance. We just take it as a baseline. The three other methods are all methods used to improve the classification tree model, as tree models are always with high variance, thus high probability to cause overfitting.
The results are listed as below:

1. Linear Discriminant Analysis

```{r Task_2_9}
lda.out<-lda(spam~.,prior=c(0.606,0.394),data=data.train)
print(lda.out)

```

Here we can see that the three attributes highly correlated with spam webmails—the frequency of “$”, “!” and “remove”, are exactly the same as what we have got from our classification tree model. What is interesting here is we can also see that the frequency of “table”, “;” and “parenthesis” shows strong negative correlation with spam webmails.

```{r Task_2_10}
pred.train<-predict(lda.out,data.train)
err(data.train$spam,pred.train$class)
pred.test<-predict(lda.out,data.test)
err(data.test$spam,pred.test$class)

```
2. Bagging

```{r Task_2_11}
# (2) bagging
set.seed(0829539)
bag.mod=randomForest(spam~.,data=data.train,classwt=c(0.606,0.394),mtry=57,ntree=5000,importance=TRUE)
bag.mod

```

```{r Task_2_12}
# (2) bagging
importance(bag.mod,plot=TRUE)
varImpPlot(bag.mod,type=2,cex=1.2)

```


```{r Task_2_13}
pred.train<-predict(bag.mod,newdata=data.train)
err(data.train$spam,pred.train)

pred.test<-predict(bag.mod,newdata=data.test)
err(data.test$spam,pred.test)

```

Actually with the MeanDecreaseGini graph we can see that the frequency of “$”, “!” and “remove” made evidently the most contribution. This is the same conclusion we have drawn from our classification tree model and Linear Discriminant Analysis. Here Out Of Bag error rate is 6.76%, which is acceptable.

3. Random Forests
```{r Task_2_14}
# (3) random forests
set.seed(0829539)
rf.mod=randomForest(spam~.,data=data.train,classwt=c(0.606,0.394),mtry=5,ntree=5000,importance=TRUE)
rf.mod

```


```{r Task_2_15}
pred.train<-predict(rf.mod,newdata=data.train)
err(data.train$spam,pred.train)

pred.test<-predict(rf.mod,newdata=data.test)
err(data.test$spam,pred.test)

```


```{r Task_2_16}
importance(rf.mod)
varImpPlot(rf.mod)

```

Random Forests Model is also based on the idea of decorrelating trees. Here we choose m=5 for the size of our predictor set. can see that although there is slightly difference, “$”, “!” and “remove” still stand for spam webmails, as well as the length of the longest uninterrupted sequence of capital letters. 


4. Gradient Boosting

Different from Bagging and Random Forests, the number of trees for Gradient Boosting is not expected to be as large as possible, since using too many trees may cause overfitting for Gradient Boosting. We start with deciding the optimal number of trees by cross-validation, for the same reason explained before.

Note that since we are going to suppose Bernoulli distribution in our gbm function, we should firstly makesure that our variable for classification is numeric type.


```{r Task_2_17}
#use distribution="bernoulli" for binary target
#interaction.depth=4, means that we fit a tree that uses 4 splits 
#(and that includes at most a 4-th order interaction)
boost.mod=gbm(spam~.,data=data.train,distribution="bernoulli",n.trees=20000,interaction.depth=4,shrinkage=0.001,cv.folds=5)
gbm.perf(boost.mod,method="cv")
#legend("topright",c("train error","CV error"),col=c("green","black"),lty=c(1,1))

```


So here we get that the optimal number of trees is 19932. We use this parameter to continue our analysis.

```{r Task_2_18}
#relative influence plot
par(cex=1.2)
summary(boost.mod,n.trees=19932)

```

We can actually get nearly the same result as we obtained from all of the former models concerning the variable most correlated to spam webmails here, that the frequency of “$”, “!” and “remove” come top.

```{r Task_2_19}
pred.train<-predict(boost.mod,n.trees=19932,newdata=data.train,type="response")
classif.train<-ifelse(pred.train>=1-pred.train,1,0)
err(data.train$spam,classif.train)
pred.test<-predict(boost.mod,n.trees=19932,newdata=data.test,type="response")
classif.test<-ifelse(pred.test>=1-pred.test,1,0)
err(data.test$spam,classif.test)

```

Then, we consider the scenario where we have different prior probabilities and unequal classification costs: C(spam|email)=10*C(email|spam). Actually the models here remain the same, and we just need to do some reclassification based on the posterior probabilities obtained from our built models and a criteria taking the classification costs as weights for posterior probabilities.


1) Linear Discriminant Analysis
```{r Task_2_20}
# (1) linear discriminant analysis
set.seed(0829539)
data.train$spam<-factor(ifelse(data.train$spam==1,"1_yes","0_no"))
data.test$spam<-factor(ifelse(data.test$spam==1,"1_yes","0_no"))
spamdata$spam<-factor(ifelse(spamdata$spam==1,"1_yes","0_no"))

lda.out<-lda(spam~.,prior=c(0.606,0.394),data=data.train)
print(lda.out)

pred.train<-predict(lda.out,data.train)
classif.train<-ifelse(1*pred.train$posterior[,2]>=10*pred.train$posterior[,1],1,0)
err(data.train$spam, classif.train)

pred.test<-predict(lda.out,data.test)
classif.test<-ifelse(1*pred.test$posterior[,2]>=10*pred.test$posterior[,1],1,0)
err(data.test$spam,classif.test)

```

2) Bagging
```{r Task_2_21}
# (2) bagging
set.seed(0829539)
bag.mod=randomForest(spam~.,data=data.train,mtry=57,ntree=5000,classwt=c(0.606,0.394),importance=TRUE)
bag.mod

#plot oob error
par(cex=1.2)
plot(1:3500,bag.mod$err.rate[1:3500,1],xlab="Number of iterations",ylab="OOB error",ylim=c(0.08,0.2),pch='',main="OOB error")
lines(1:3500,bag.mod$err.rate[1:3500,1],col="red")
#plot variable importance
importance(bag.mod,plot=TRUE)
varImpPlot(bag.mod,type=2,cex=1.2)

pred.train<-predict(bag.mod,type="prob",newdata=data.train)
classif.train<-ifelse(1*pred.train[,2]>=10*pred.train[,1],1,0)
err(data.train$spam,classif.train)
pred.test<-predict(bag.mod,type="prob",newdata=data.test)
classif.test<-ifelse(1*pred.test[,2]>=10*pred.test[,1],1,0)
err(data.test$spam,classif.test)

```

3) Random Forests
```{r Task_2_22}
# (3) random forests
set.seed(0829539)
rf.mod=randomForest(spam~.,data=data.train,mtry=5,ntree=5000,classwt=c(0.606,0.394),importance=TRUE)
rf.mod

pred.train<-predict(rf.mod,type="prob",newdata=data.train)
classif.train<-ifelse(1*pred.train[,2]>=10*pred.train[,1],1,0)
err(data.train$spam,classif.train)
pred.test<-predict(rf.mod,type="prob",newdata=data.test)
classif.test<-ifelse(1*pred.test[,2]>=10*pred.test[,1],1,0)
err(data.test$spam,classif.test)

#variable importance plot
#two measures are reported: 
#increase in MSE computed on OOB samples when removing variable
#decrease in node impurity (or increase in node purity) resulting from splits on the variable
importance(rf.mod)
varImpPlot(rf.mod)

```

4) Gradient Boosting
```{r Task_2_23}
# (4) gradient boosting
set.seed(0829539)
data.train$spam<-ifelse(data.train$spam=="1_yes",1,0)
data.test$spam<-ifelse(data.test$spam=="1_yes",1,0)
spamdata$spam<-ifelse(spamdata$spam=="1_yes",1,0)

#use distribution="bernoulli" for binary target
#interaction.depth=4, means that we fit a tree that uses 4 splits 
#(and that includes at most a 4-th order interaction)
boost.mod=gbm(spam~.,data=data.train,distribution="bernoulli",n.trees=20000,interaction.depth=4,shrinkage=0.001,cv.folds=5)
gbm.perf(boost.mod,method="cv")
legend("topright",c("train error","CV error"),col=c("green","black"),lty=c(1,1))

#relative influence plot
par(cex=1.2)
summary(boost.mod,n.trees=19932)

pred.train<-predict(boost.mod,n.trees=19932,newdata=data.train,type="response")
classif.train<-ifelse(1*pred.train>=10*(1-pred.train),1,0)
err(data.train$spam,classif.train)
pred.test<-predict(boost.mod,n.trees=19932,newdata=data.test,type="response")
classif.test<-ifelse(1*pred.test>=10*(1-pred.test),1,0)
err(data.test$spam,classif.test)

```
We arrange all the training errors and test errors into a whole table below, and calculate the sensitivities as well as the false positive rates for the test set, using the probability tables we printed out:

(image)

We can see that as we claimed at the beginning, Linear Discriminant Analysis performs the worst in both scenarios, Random Forests performs the best in scenario a), where we have different prior probabilities but equal classification costs, and Gradient Boosting performs the best in scenario b), where we have different prior probabilities and unequal classification costs. Generally, for scenario a), the performance of Random Forests and Gradient Boosting is equally good, and Bagging also performs good, just with slightly higher test error than Random Forests and Gradient Boosting. However, for scenario b), the accuracy of Random Forests decreases much faster than Bagging. 

What’s more, as is expected, in scenario b), since missclassifying a not spam webmail into spam ones costs much more than missclassifying a spam webmail into not spam ones, our models all tend to classify more cases as not spam ones to minimize the cost. As a result, the false positive rate is very low or all models, but the accuracy and sensitivity decreases significantly for all models, since a number of real positive cases are classified as not spam webmails. Gradient Boosting and Bagging are relatively more stable here compared with the other two models.

