---
title: "Task 5"
output: 
  pdf_document: 
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup5, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 5
```{r Task_5_setup, include=FALSE}
base_url <- '~/GitHub/kul-multivariate-a2/src/'
load(paste0(base_url,"confusion.Rdata"))
#install.packages("smacof")
library(smacof)
library(dplyr)
library(qdapRegex)
library(stringr)
library(kableExtra)
#write.csv(confusion,"~/GitHub/kul-multivariate-a2/src/cnf.csv", row.names = FALSE)
alphabet <- c(letters,1:9,0)
translate <- confusion
colnames(translate) <-alphabet
rownames(translate) <-alphabet

sim <- (confusion+t(confusion))/2 #similarity
dissim <- 100-sim #dissimilarity
for (i in 1:36){
  dissim[i,i]<-0   #diagonal 0
} 

tsim <- (translate+t(translate))/2 #similarity
tdissim <- 100-tsim #dissimilarity
for (i in 1:36){
  tdissim[i,i]<-0   #diagonal 0
} 

```

## Introduction

  In a study on Morse code signals Rothkopf (1957) asked 598 subjects to indicate for each pair of signals (acoustically presented after each other) whether the two signals were equal.
  Note that each signal consists of a sequence of short beeps (0.05 sec) and long beeps (0.15 sec) separated by a silence (0.05 sec).
  The data set confusion.Rdata contains a 36 x 36 matrix of confusion rates for all pairs of signals included in the study. The signals in the subsequent rows of the confusing matrix are the letters of the alphabet A, B, C… in alphabetical order and the numbers 1,2,3,4,5,6,7,8,9,0.
  The value in cell (i,j) of the matrix represents the percentage of subjects who indicated that signal i was equal to signal j, when signal j was presented after signal i.
 
 
## Methodology

  First of all and in order to observe dissimilarities, smacofSym() will be used to conduct a MDS with 2 dimensions. 
  
  After 

  We compute a symmetric matrix of similarities by taking the average of corresponding elements in the confusion matrix above and below the diagonal, 
next we convert the matrix of similarities into a matrix of dissimilarities, and finally we put all diagonal elements of the dissimilarity matrix equal to 0:
..................
## Results
  ..................

```{r Task_5_0, include=FALSE}
sim <- (confusion+t(confusion))/2 #similarity
dissim <- 100-sim #dissimilarity
for (i in 1:36){
  dissim[i,i]<-0   #diagonal 0
} 

str(sim) 
```

### Question 1
1. Use *smacofSym()* to conduct MDS with 2 dimensions and assuming different measurement levels (i.e., _ratio, interval, mspline, ordinal_) for the observed dissimilarities.

#### 1.1 Getting information from dissimilarity matrix. 
  Here ratio, interval, mspline and ordinal values are obtained.
    
```{r Task_5_1N, include=FALSE, eval=TRUE}
#ratio
tm1 <- smacofSym(delta=tdissim, ndim=2, type="ratio", init="torgerson")
#interval
tm2 <- smacofSym(delta=tdissim, ndim=2, type="interval", init="torgerson")
#mspline
tm3 <- smacofSym(delta=tdissim, ndim=2, type="mspline", spline.degree =4, 
                spline.intKnots = 4, init="torgerson")
#ordinal
tm4 <- smacofSym(delta=tdissim, ndim=2, type="ordinal", init="torgerson")
```
```{r Task_5_1Y1, include=TRUE, eval=TRUE}
#ratio
m1 <- smacofSym(delta=dissim, ndim=2, type="ratio", init="torgerson")
#interval
m2 <- smacofSym(delta=dissim, ndim=2, type="interval", init="torgerson")
#mspline
m3 <- smacofSym(delta=dissim, ndim=2, type="mspline", spline.degree =4, 
                spline.intKnots = 4, init="torgerson")
#ordinal
m4 <- smacofSym(delta=dissim, ndim=2, type="ordinal", init="torgerson")
```
```{r Task_5_1Y2, include=TRUE, eval=TRUE}
#ratio
mmmm1 <- smacofIndDiff(tdissim, type= "ratio", constraint = "indscal", ties="primary")
#interval
m2 <- smacofSym(delta=dissim, ndim=2, type="interval", init="torgerson")
#mspline
m3 <- smacofSym(delta=dissim, ndim=2, type="mspline", spline.degree =4, 
                spline.intKnots = 4, init="torgerson")
#ordinal
m4 <- smacofSym(delta=dissim, ndim=2, type="ordinal", init="torgerson")
```
```{r Task_5_2, include=FALSE, eval=TRUE}
## jUST df creation to be used in kable
dt <- data.frame(
  "m1" = round(c(m1$stress),3),
  "m2" = round(c(m2$stress),3),
  "m3" = round(c(m3$stress),3),
  "m4" = round(c(m4$stress),3)
)
```
```{r Task_5_2n, include=FALSE, eval=TRUE}
## jUST df creation to be used in kable
tdt <- data.frame(
  "tm1" = round(c(tm1$stress),3),
  "tm2" = round(c(tm2$stress),3),
  "tm3" = round(c(tm3$stress),3),
  "tm4" = round(c(tm4$stress),3)
)
```


```{r Task_5_4, include=TRUE, echo=FALSE}
dt %>%
  kbl(caption = "Stress-1  table values") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```

```{r Task_5_4n, include=FALSE, echo=FALSE}
tdt %>%
  kbl(caption = "Stress-1  table values") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```
```{r Task_5_5N, include=TRUE, echo=FALSE}

par(mfrow=c(2,2))
plot(tm3,plot.type="resplot",main="residual plot ordinal MDS")
plot(tm3,plot.type="Shepard",main="Shepard diagram ordinal MDS")
plot(tm4,plot.type="resplot",main="residual plot spline MDS")
plot(tm4,plot.type="Shepard",main="Shepard diagram spline MDS")
#configuration ordinal MDS

```
```{r Task_5_5, include=TRUE, echo=FALSE}

par(mfrow=c(2,2))
plot(m3,plot.type="resplot",main="residual plot ordinal MDS")
plot(m3,plot.type="Shepard",main="Shepard diagram ordinal MDS")
plot(m4,plot.type="resplot",main="residual plot spline MDS")
plot(m4,plot.type="Shepard",main="Shepard diagram spline MDS")
#configuration ordinal MDS

```
```{r Task_5_6n,include=TRUE,  include=TRUE}
plot(tm3,plot.type="conf")
```
```{r Task_5_6, include=TRUE}
plot(m3,plot.type="conf")
```

### Question 2
2. Evaluate the goodness of fit of solutions with different measurement levels using _stress-1_, and by computing stress norms with the functions _randomstress()_ and _permutation()_. Discuss which solution you would select. Investigate the stability of the selected solution using the _Jackknife_.

```{r Task_5_7, include=FALSE, eval=FALSE}
#stress norm
#set.seed(1)
#rstress<-randomstress(n=36,ndim=2,nrep=500,type="ordinal")
#distribution of stress for random data
#mean(rstress)-2*sd(rstress)
#permutation test
#set.seed(1)
#perm.car<-permtest(m3,nrep=500)
#plot distribution stress
#par(mfrow=c(1,2),pty="s")
#hist(rstress,main="stress random data")
#hist(perm.car$stressvec,main="stress permuted data")
#stability of solution using jackknife
#jack.car<-jackmds(m3)
#plot(jack.car,xlim=c(-1.2,1.2),ylim=c(-1,1))
```

### Question 3

3. Construct a data set of external variables that describe the signals (e.g. length of the signal, proportion of short beeps in the signal, etc.). Use an _MDS biplot_ to project the external variables in the configuration plot of the selected solution, and interpret the results of the analysis. 

```{r Task_5_8N, include=TRUE}
v_signal <- colnames(confusion)
v_signal <- str_replace(v_signal, " ", "")
v_signal <- str_replace(v_signal, "- -", "--")
v_signal <- str_replace(v_signal, "· -", "·-")
v_signal <- str_replace(v_signal, "· ·", "··")
v_signal <- str_replace(v_signal, "- ·", "-·")
alphabet <- c(letters,1:9,0)
another_matrix <- data.frame(v_signal) %>% mutate(alph = alphabet, 
         size = nchar(trimws(v_signal)), 
         ndots =  str_count(v_signal, "·"),
         nlines = str_count(v_signal, "-"),
         sig_s_length = (str_count(v_signal, "·")*0.05)+(str_count(v_signal,"-")*0.15),
         t_s_length = (str_count(v_signal, "·")*0.05) + 
              (str_count(v_signal,"-")*0.15) +
           ((nchar(trimws(v_signal))-1 )*0.05), 
         dots_s_length = (str_count(v_signal,"·")*0.05),
         lines_s_length = (str_count(v_signal,"-")*0.15)) %>% print()
```
```{r Task_5_8Y, include=TRUE}

#bimorse <- biplotmds(sim, another_matrix)

#plotMDS(dissim)

#fit two-dimensional indscal model assuming different measurement levels
#m1<-smacofIndDiff(dissim, type="ratio", constraint="indscal",ties="primary")
#m2<-smacofIndDiff(dissim, type="interval", constraint="indscal",ties="primary")
#m3<-smacofIndDiff(dissim, type="ordinal", constraint="indscal",ties="primary")
#round(c(m1$stress,m2$stress,m3$stress),3)

#res <- mds(dissim)
#fitbi <- biplotmds(res, morsescales[,2:3])
#plot(fitbi, main = "MDS Biplot", vecscale = 0.5)
#alphabet <- c(letters,1:9,0)

#confusion2 <- confusion
#colnames(confusion2) <- paste(colnames(confusion), "=", c(letters,1:9,0)) 
#rownames(confusion)

#compute MDS biplot: run multivariate linear regression of
#external variables on configuration

#fitFace <- mds(dissim, type = "ordinal")

#biFace <- biplotmds(fitFace, another_matrix[,-c(1:2)])

#plot(biFace, main = "Biplot Vector Representation", 
#     vecscale = 0.8, xlim = c(-1.5, 1.5), 
#     vec.conf = list(col = "brown"), pch = 20, cex = 0.5)

```